---
title: "Geochem pipeline - minimal example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup_libraries, message=FALSE, warning=FALSE}
library(geochem)
library(rio)
library(dplyr)
library(ggplot2)
```

```{r setup_theme}
theme_glab = function(
  base_size=12,
  base_family="",
  base_line_size=base_size / 180,
  base_rect_size=base_size / 180
) {

  # Assign font family up front
  font = "Helvetica"

  ggplot2::theme_bw(
    base_size=base_size, 
    base_family=base_family,
    base_line_size=base_line_size
  ) %+replace%
  ggplot2::theme(
    legend.background=element_blank(),
    legend.title=element_text(
      color=rgb(100, 100, 100, maxColorValue=255),
      size=rel(0.65),
      hjust=0
    ),
    legend.text=element_text(
      color=rgb(100, 100, 100, maxColorValue=255),
      size=rel(0.65)
    ),
    legend.key.size=unit(0.8, "lines"),
    plot.title=element_text(
      color=rgb(100, 100, 100, maxColorValue=255), 
      hjust=0
    ),
    axis.title=element_text(
      color=rgb(100, 100, 100, maxColorValue=255),
      size=rel(0.65)
    ),
    axis.text=element_text(
      color=rgb(100, 100, 100, maxColorValue=255),
      size=rel(0.65)
    ),
    plot.caption=element_text(
      color=rgb(100, 100, 100, maxColorValue=255),
      size=rel(0.35),
      hjust=1
    ), 
    panel.grid.major=element_blank(),   
    panel.grid.minor=element_blank(),   
    panel.border=element_rect(
      colour=rgb(100, 100, 100, maxColorValue=255),
      fill=NA
    ),
    complete=TRUE
  )
}
```


<br>


## Data folder

Here you define where your data folder is located:

```{r setup_variables}
PROJECT_NAME = "PROJECT_NAME"
DATA_DIR   = file.path("../data", PROJECT_NAME)
FIGURES_DIR  = file.path("../data", PROJECT_NAME, "geochem-figures")
```

We strongly recommend the following folder structure, where:

* `PROJECT_NAME` is the name of your project
* The `icp-ms` folder should only contain the `PROJECT_NAME.xlsx` file **before** running the pipeline
* The `geochem-figures` folder should be empty **before** running the pipeline

```
data
└── PROJECT_NAME
    ├── PROJECT_NAME_env_dataset.xlsx
    ├── icp-ms
    │   ├── PROJECT_NAME.xlsx
    └── geochem-figures
```


<br>


## Load project's `env_dataset`

The `env_dataset` will be used for plotting both DIC and IC data.
However, ICP-MS data should be carefully curated (see ICP-MS section).

```{r data_load, message=FALSE}
data_df = rio::import(
  file=file.path(
  DATA_DIR,
  paste0(PROJECT_NAME, "_env_dataset.xlsx")
  ),
  sheet="envdata_template",
  skip=1
)

# Remove first row containing the units
data_df = data_df %>% slice(2:nrow(data_df))

# Convert data columns to numeric
data_df = type.convert(data_df, as.is=TRUE)

# Convert SiteID to ordered factor for plotting
data_df = data_df %>% mutate(SiteID=factor(SiteID, levels=data_df$SiteID))

head(data_df %>% select(SiteID, ExpID, year))
```


<br>


## Geochemistry analysis


### IRMS analysis

In this section, data from IRMS will be used to analyse the content of dissolved inorganic carbon (DIC).

```{r figure_dic}
figure_dic = ggplot(
    data=data_df,
    aes(
      x=SiteID,
      y=d13DIC
    )
  ) +
  geom_boxplot(fill="lightgray") +
  scale_x_discrete(limits=unique(data_df$SiteID)) +
  theme_glab() +
  # Rotate X axis labels
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))

ggsave(
  plot=figure_dic,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-dic.svg")
  ),
  width=7,
  height=7
)

figure_dic
```


<br>


### IC analysis

In this section, data from IC will be used to analyse the content of anions and cations.


#### Preprocessing

Before running the pipeline, we need to preprocess the values in the `env_dataset`:

```{r process_ic}
# Perform unit conversions and calculate values for the different plots
ic_df = process_ic(data_df)

head(ic_df %>% select(SiteID, HCO3.meq, Cl.meq, SO4.meq, Na.meq, K.meq, IB))
```


#### Figure: major anions

```{r figure_major_anions, warning=FALSE}
figure_major_anions = plot_base_major_anions() +
  geom_point(
    data=ic_df,
    aes(
      x=SO4.meq,
      y=Cl.meq,
      z=HCO3.meq,
      fill=SiteID
    ),
    size=4.5,
    shape=21,
    stroke=0.3
  ) +
  theme_glab()

ggsave(
  plot=figure_major_anions,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-major-anions.svg")
  ),
  width=7,
  height=7
)

figure_major_anions
```


#### Figure: major cations

```{r figure_major_cations, warning=FALSE}
figure_major_cations = plot_base_major_cations() +
  geom_point(
    data=ic_df,
    aes(
      x=Ca.meq,
      y=Mg.meq,
      z=Na.meq + K.meq,
      fill=SiteID
    ),
    size=4.5,
    shape=21,
    stroke=0.3
  ) +
  theme_glab()

ggsave(
  plot=figure_major_cations,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-major-cations.svg")
  ),
  width=7,
  height=7
)

figure_major_cations
```


#### Figure: water maturity (Giggenbach)

```{r figure_maturity_gigg, warning=FALSE}
figure_maturity_gigg = plot_base_water_maturity(type="giggenbach") +
  geom_point(
    data=ic_df,
    aes(
      x=k/100,
      y=na/1000,
      z=sqrt(mg),
      fill=SiteID
    ),
    size=4.5,
    shape=21,
    stroke=0.3
  ) +
  theme_glab()

ggsave(
  plot=figure_maturity_gigg,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-water-maturity-giggenbach.svg")
  ),
  width=7,
  height=7
)

figure_maturity_gigg
```


#### Figure: water maturity (Duchi)

```{r figure_maturity_duchi, warning=FALSE}
figure_maturity_duchi = plot_base_water_maturity(type="duchi") +
  geom_point(
    data=ic_df,
    aes(
      x=k/10,
      y=na/400,
      z=sqrt(mg),
      fill=SiteID
    ),
    size=4.5,
    shape=21,
    stroke=0.3
  ) +
  theme_glab()

ggsave(
  plot=figure_maturity_duchi,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-water-maturity-duchi.svg")
  ),
  width=7,
  height=7
)

figure_maturity_duchi
```


#### Figure: Langelier-Ludwig diagram

```{r figure_ll}
figure_ll = plot_base_ll() +
  geom_point(
    data=ic_df,
    aes(
      x=r_bicarb,
      y=r_na_k,
      fill=SiteID
    ),
    size=4.5,
    shape=21,
    stroke=0.3
  )

ggsave(
  plot=figure_ll,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-langelier-ludwig.svg")
  ),
  width=7,
  height=7
)

figure_ll
```


#### Figure: Gibbs diagram

```{r figure_gibbs}
figure_gibbs = plot_base_gibbs() +
    geom_point(
        data=ic_df,
        aes(
            x=na/(na+ca),
            y=tds,
            fill=temp
        ),
        size=4.5,
        shape=21,
        stroke=0.3
    ) +
    scale_fill_viridis_c(option="inferno", limits=c(0, 122)) +
    geom_label(
        data=ic_df,
        aes(
            x=na/(na+ca),
            y=tds,
            label=SiteID,
            color=SiteID
        ),
        nudge_y=0.25
    )

ggsave(
    plot=figure_gibbs,
    filename=file.path(
        FIGURES_DIR,
        paste0(PROJECT_NAME, "-gibbs.svg")
    ),
    width=7,
    height=7
)

figure_gibbs
```


#### Figure: Piper diagram

```{r figure_piper}
# Project data on the Piper plot
piper_data = transform_data_piper(
    Mg=ic_df$Mg.meq.perc,
    Ca=ic_df$Ca.meq.perc,
    Cl=ic_df$Cl.meq.perc,
    SO4=ic_df$SO4.meq.perc,
    id=ic_df$SiteID
)
piper_data = piper_data %>% rename(SiteID=id)

figure_piper = plot_base_piper() +
    geom_point(
        data=piper_data,
        aes(
            x=x,
            y=y,
            fill=SiteID
        ),
        size=4.5,
        shape=21,
        stroke=0.3
    )

ggsave(
    plot=figure_piper,
    filename=file.path(
        FIGURES_DIR,
        paste0(PROJECT_NAME, "-piper.svg")
    ),
    width=7,
    height=7
)

figure_piper
```


<br>


### ICP-MS analysis

The raw data from the instrument must be processed for easier examination.
The selection of reasonable measurements can be performed manually or automatically.

To perform a **manual selection**, run the next chunk and examine the generated file:
`PROJECT_NAME_check.xlsx`

To perform an **automatic selection**, run the following line:

`icp_df = icp_df %>% dplyr::filter(value_sd_check != "DISCARD")`


```{r setup_variables_icp}
# NOTE: set these two variables
ICP_FILENAME = "PROJECT_NAME.xlsx"
ICP_BLANK_NAME = "BLANK"

# NOTE: automatically set by the package
ICP_DIR = file.path(DATA_DIR, "icp-ms")
```


#### Load and preprocess the data to a readable format

```{r process_icp, warning=FALSE, message=FALSE}
icp_df_check = process_icp(
  filepath=file.path(ICP_DIR, ICP_FILENAME),
  blank_name=ICP_BLANK_NAME
)

write.csv(
  x=icp_df_check,
  file=file.path(
    ICP_DIR,
    paste0(tools::file_path_sans_ext(ICP_FILENAME), "_check.csv")
  ),
  row.names=FALSE
)

dim(icp_df_check)
```


#### Perform the selection

```{r select_icp}
# NOTE: icp_df is the selected data, whether manually or automatically selected

icp_df = select_icp_auto(
  df=icp_df_check,
  blank_name=ICP_BLANK_NAME
)

# icp_df = icp_df_check

dim(icp_df)
```


#### Filter metals

```{r filter_metals}
# Remove selected elements
icp_df = icp_df %>%
  dplyr::filter(
    element != "Li",
    !(element == "Na" & gas == "No Gas"),
    element != "Mg",
    !(element == "P" & gas == "No Gas"),
    !(element == "S" & gas == "No Gas"),
    element != "K",
    element != "Ca",
    !(element == "Si" & gas == "No Gas")
  )
```


#### Figure: metals

```{r metal_color_mapping}
# Set the color for each metal
selected_metals = read.csv(
    text="element,color
      Fe,#f57900
      V,#4e9a06
      Mn,#563566
      Co,#204a87
      Ni,#555753
      Cu,#8ae234
      Zn,#babdb6
      Se,#edd400
      Mo,#a40000
      W,#8f5902"
  ) %>%
  # Replace indentation whitespaces
  mutate(
    element=stringr::str_replace(
      string=element,
      pattern="\\s+",
      replacement=""
    )
  )

# Create a named vector
selected_metals = selected_metals %>% tibble::deframe()

icp_df_selected = icp_df %>%
  dplyr::filter(element %in% names(selected_metals))
```

```{r figure_metals}
figure_metals = plot_metals(icp_df_selected)

ggsave(
  plot=figure_metals,
  filename=file.path(
    FIGURES_DIR,
    paste0(PROJECT_NAME, "-metals.svg")
  ),
  width=7,
  height=7
)

figure_metals
```


#### Integration with the `env_dataset`

The next chunk of code saves the data in wide format, ready to be copied into the `env_dataset`:

```{r icp_env_dataset}
# Save processed ICP-MS data in wide format
icp_df_wide = icp_df %>%
  ungroup() %>%
  select(sample, element, isotope, dilution, gas, concentration) %>%
  tidyr::pivot_wider(
    id_cols=c(
      sample,
      dilution,
      gas
    ),
    names_from=c(
      isotope,
      element
    ),
    names_sep="_",
    values_from=concentration
  )

write.csv(
  x=icp_df_wide,
  file=file.path(
    ICP_DIR,
    paste0(tools::file_path_sans_ext(ICP_FILENAME), "_env_dataset_icp.csv")
  ),
  row.names=FALSE
)
```
